---
title: "Data analysis"
author: "Banele Ngcethane"
date: "2023-10-04"
output: html_document
---


## shareprice  

```{r}
library(readxl)
library(forecast)
library(zoo)

View(sharepricedata_2023_)

time_series_data=sharepricedata_2023_
time_series_graph=ts(time_series_data$Value,
                     frequency = 12, start = c(1960,1),end=c(2022,12))

```

## Including Plots


```{r pressure, echo=FALSE}
plot(decompose(time_series_graph))
ggtsdisplay(time_series_graph)
acf(time_series_graph)
pacf(time_series_graph)
```
TEST FOR STATIONARITY  
H0: data has a unit root and is not stationary  
H1: data has no unit root and is stationary  
at 5%, if the p-value > 0.05 the data is not stationary
```{R}
library(aTSA)
library(tseries)
adf.test(time_series_graph)

```
  
## differenced data  

```{r}
first_differenced_data=diff(time_series_graph)
plot(first_differenced_data,main="first differenced data", 
     xlab="Year",Ylab="Differenced Values",col="black")
acf(first_differenced_data,lag.max = 100)
pacf(first_differenced_data)
ggtsdisplay(first_differenced_data)

adf.test(first_differenced_data)
kpss.test(time_series_graph)

ndiffs(time_series_graph)

second_differenced_data = diff(time_series_graph,differences = 1)
plot(second_differenced_data,main="Second differenced data",
     xlab="year",ylab="Second differenced values",col="black")
ggtsdisplay(second_differenced_data)
acf(second_differenced_data)
pacf(second_differenced_data)
adf.test(second_differenced_data)
```

   
The ACF and the PACf of the second diferenced data suggest a  
SARIMA(6,2,1)(0,0,1)

## LOG DATA  
```{r}
log_data=sqrt(time_series_graph)
plot(log_data,main="transformed data (log)",
     ylab="log data",xlab="year",col="black")
acf(log_data)
pacf(log_data)
ggtsdisplay(log_data)
adf.test(log_data)


#log differenced data
first_differenced_log_data = diff(log_data)
plot(first_differenced_log_data,main="first differenced log data",
     ylab="differenced log values",xlab="year",col="black")
acf(first_differenced_log_data)
pacf(first_differenced_log_data)
ggtsdisplay(first_differenced_log_data)
adf.test(first_differenced_log_data)

second_differenced_log_data = diff(log_data,differences = 2)
plot(second_differenced_log_data,main="second differenced log data",
     xlab="year",ylab="second differenced log values",col="black")
acf(second_differenced_log_data)
pacf(second_differenced_log_data)
ggtsdisplay(second_differenced_log_data)
adf.test(second_differenced_log_data)

```
   
the log data is non stationary since the p-value is not less than 0.05  
the second differenced data is stationary  
and suggest a SARIMA(5,2,0)(0,0,1)  
## models  
model by non transformed and transformed data  

```{r}
library(forecast)
model1=forecast::Arima(time_series_graph,order=c(6,2,1),seasonal = c(0,0,1))
model2=forecast::Arima(time_series_graph,c(5,2,0),seasonal = c(0,0,1))
plot(time_series_graph)
lines(model1$fitted,col="red")
lines(model2$fitted,col="blue")


plot(model1$fitted,col="red")
plot(model2$fitted,col="blue")

```
   
#accuracy of the models
```{r}
accuracy(model1)
accuracy(model2)
model1$aic
model1$bic
model2$aic
model2$bic

```

